Formalizing Abstract Rewriting

### 2025-02-22 Relations/notes.tex
AP notes

\input{macros}

@@

This document aims to bring together the various notes from report, rewriting qs, and counterexamples.

\section{Todo}
\subsection*{Week starting 11th November}
\begin{enumerate}
  \item In the report, give a classical proof that the $Inc$ Property
  implies that no $R$-increasing sequence is bounded.

  Conclude that every relation satisfying $Inc$ also satisfies $RP$.

  Therefore, the proof of Theorem 1.2.3(iii) using $RP$ instead of $Inc$
  is generalizing the claim of the theorem (as stated in [TeReSe]).
  \item Complete the proof in ARS.agda of Theorem 1.2.3.(iii)
  \item Organize notes.tex into three sections: TODO, Open Qs, What we've proved. Make sure all notes from the other
  tex files are sorted here correctly.
  \item Tidy up and standardize the notation used in ars.agda and other agda files 
  \item Carry on working through report.tex.  
  \item 
\end{enumerate}
\subsection*{General todo}
\begin{enumerate}

  \item  Make the first argument to "being a normal form" implicit?
  \item Define WNFP to be:
  \[\forall a b c \to b \in NF \to a R^* b \to a R^* c \to c R^* b \]
  \item Does WNFP and WN imply CR?  Looks like it!
  \item  Another interesting one? $WN^*$: hereditarily $WN$ ($WN$ for the reduction graph starting from a given element $x$.)

  Could even consider $SN^*$, $CR^*$ etc.
  \item Define: Inc
  \item Define: Ind
  \item Define: $\mathrm{CR}^{\le 1}$
  \item capture all classical properties/asssumptions we might want to use and justify their use. 
  \item Principles we are considering
  \begin{itemize}
    \item Decidability of equality on $A$
    \item Decidability of the relation $R$
    \item When relevant, decidability of the given precicate $P$
    \item Not-not-closure of being accessible
    \item Not-not-closure of $R$
    \item Not-not-closure of $P$
    \item Decidability of whether a given element is an $R$-normal form
    \item When relevant, local/element-wise versions of the above
    \item Markov's principle:
    \begin{itemize}
      \item In $wfMin_0 \to wfSeq$.

      Given a sequence $s : \nat \to A$, let $E(a) = \exists n \in \nat. s n \equiv a$.

      Need: $\forall a. \lnot\lnot E(a) \to E(a)$, i.e.,
      \[\lnot\lnot \big(\exists n \in \nat. s n \equiv a\big)
            \to \big(\exists n \in \nat. s n \equiv a\big)\]

    \end{itemize}
  \end{itemize}
  \item Write a guide to the code
  \item More laws about closure operations: Monotonicity, idempotency

\end{enumerate}
\newpage
\section{Open Questions }
\begin{enumerate}
  \item
     \[ WNg \land UN \to CRelem :
     \forall (R : \mathscr{R} A) \to WN R \to \forall x \to is R -UN x \to is R -CR x \]
     \item $WN R \to UN R \to \omega-bdd R \to SN R$ (1.2.3.ii-)
      \item Finish formulating the ``compactness property".  Two candidates:
       \item ``Every cocone for a given infinite sequence loops back to some point in the sequence"
       \item ``If an $R^*$-sequence has a cocone then it's constant after some point on.''
       \item is $\omega$-bounded implied by WN and the following confluence property:
       $a \to^* n \in NF$, and $a \to b$ then $b \to^* n $.
  \item (2024.10.10) Is being $R$-WN inductive?
  \item (2024.10.10) Does $R$ being finitely branching imply
   that, for an inductive $\phi$, if $\lnot \phi(x)$ then
   $\exists y. Rxy \land \lnot \phi(y)$?

  \item If WCR holds, is WN$\downarrow\subseteq$WN ?
  \item Same question for next bullet in 2.2
  \item Same for bullet 3 in 2.2 (but pointwise!)
  \item \item Look at restricting WFmin to $\lnot\lnot$-closed predicates
  \item Status of important hypotheses:
    \begin{itemize}
      \item Deciding whether a given element $x$ reduces to a given element $y$
      (Is $R$ itself decidable?)
      \item Deciding whether a given element $x$ reduces to some element $y$
      (Is being an $R$-normal form decidable?)

      {\textcolor{red}{This is needed to show that SN $\Rightarrow$ WN.}}

      \item If a given element $x$ is \emph{not} a normal form,
      exhibiting some element $y$ that it reduces to.
      (This is related to DeMorgan/Markov/etc. properties that came up in WF file.)

      \item Does UN-lemma \emph{require} the decidability assumption?
    \end{itemize}
  \item META questions: Which of the ARS theorems can be made ``local/pointwise'',
  for example,
  \begin{itemize}
    \item NL+: $WCR(R) \land SN(x) \to CR(x)$?  (Yes?)
    \item NL++: $WCR(x) \land SN(x) \to CR(x)$? (No.)
  \end{itemize}
  \item Can "sequential" properties like RP, RP-, isWFseq, isWFseq-, $\omega$-bounded, etc.,
    be related to the cofinality property.
    \item What happens if properties involving sequences are reformulated for
  $R^r$? (Cannot decide whether the given sequence is finite or infinite.)
  \item Knaster-Tarski Lemma: Most general form? Alternative formulations?
  \item Knaster--Tarski in a well-founded setting: what additional
  hypotheses are necessary?
\end{enumerate}
\subsection*{well-foundedness}
\begin{enumerate}
  \item For a given element $x$, is the property of
  being accessible $\lnot\lnot$-closed? I.E.,
  Does $x$ being $\lnot\lnot$-accessible imply that $x$ is accessible?
  (Conjecture: No.)

  \item As a special case, does being weakly (accessible) well-founded imply being $\lnot\lnot$-wellfounded?
  (Conjecture: No.)

  Also: Same question about inductive notion of well-foundedness.

  \item ($\star$) Given a well-founded relation, does every non-empty subset
  have a minimal element?

  \emph{Problem.} Need to decide whether, for a given $x \in U$,
   the set $\{y | Ryx\}$ is empty.

   \item IF every non-empty subset has a minimal element, does this imply
   either of the weak forms of well-foundedness: $\mathtt{isWFacc-}$ or
   $\mathtt{isWFind-}$ ?

   \item Does $\mathtt{isWFmin-}$ imply $\mathtt{isWFacc-}$ or $\mathtt{isWFind-}$?

   \emph{Problem.} Need to go from $\lnot (\forall y. R y d \to \phi y)$
   and $\forall y. R y d \to \lnot \lnot \phi y$ to $\bot$.

   In terms of accessibility, it should suffices to assume accessibility is
   $\lnot\lnot$-closed, since being related to y is not relevant for that. (?)

   \item Does sequential well-foundedness (no decreasing sequence) imply
   any of the other notions, e.g., \texttt{WFmin-} ?

   \emph{Note.}  This seems to require the most classical assumptions:
   $\lnot\lnot$-closure of $\phi$, relativized De Morgan law,
   Markov's principle, etc.
   \textsc{Remark.}
   $\lnot\lnot$-closure of accessibility should solve problems 1 and 4.

\end{enumerate}

\newpage 
\section{What has been proven}
\begin{itemize}
  \item \sacomment{List all the constructive proofs we've put together!}
  \item SN implies bounded
  \item RP- implies RP
  \item The infinite haircomb refutes $WN \land RP \to \omega$-bounded.  But it is not WCR.
  \item We have formalized proposition 1.1.9, 1.1.10, 1.1.11 from [TeReSe]. 
  \item We have a proof for Theorem 1.2.1 (Newlman's Lemma) (though we have just one proof of multiple possible proofs).
  \item We have a proof for all parts of Theorem 1.2.2
  \item We have a proof for Theorem 1.2.3 i, iv. 
  \item Reduction closure properties for SN, NF, UN, recurrent property.

  So, does $WN, RP, WCR$ imply $\omega$-bounded?
  Answer: yes, that is what we have proved classically.
\end{itemize}

\newpage

@@

>>>>

\usepackage{xcolor}

\newcommand{\hired}[1]{{\color{red}{#1}}}

@@

\section*{Notes}

\section{Plan}
\begin{itemize}
  \item Start by discussing motivation
  \item Then contributions
  \item Plan of the paper
  \item At some point, there should be like a summary of what's been done and what are some interesting discoveries
\end{itemize}

In the conclusion, we should recap the main takeaways from each piece.
The key role of finitely branching condition in making many results effective.
That it also collapses many well-foundedness properties.
The ``sufficiency frontier'' for completeness (termination and confluence).

Also discuss open problems, unprovable implications, etc.
Maybe mention the Kleene tree idea, or at least include the reference to the Bezem-Coquand paper.

\section{Motivation}

The standard presentation in Terese makes liberal use of classical logic.
Should stress effectivity of type-theoretic formalization.
\section{Contributions}

\begin{itemize}
  \item Formalization of ``elementary ARS theory'' as presented in
  Chapter 1 of [TeReSe].
  \item An ontology of termination and confluence properties, and a detailed analysis of logical relations between them.
  \item Defining several new ARS concepts that serve to streamline and generalize a number of relationships.
  \item With the above, obtain marginal improvements to classical confluence and termination criteria
  \item Collecting implications between constructive notions of well-foundedness
\end{itemize}

\section{Formalization principles}

\begin{itemize}
  \item No function extensionality
  \item No univalence, uniqueness of identity proofs, cubical, or any other
  assumptions related to equality
  \item Minimize the use of classical logic; every place where it was necessary,
  make decidability hypotheses explicit in every place where it could not be avoided.
  \item The vast majority of results are completely constructive and therefore,
  they compute.
  \item The goal is to have the results of basic abstract rewriting
  formalized in a canonical way, so that they can be used in building
  libraries of formalized programming language theory
  \item Since this application ....
  \item In other words, we wanted to stay as close as possible to the spirit of type theory based on the Curry--Howard isomorphism, so that in every explicitly given situation, our proofs will compute.
\end{itemize}

One big item in this development was studying various incaranations of the concept of well-foundedness in constructive context.
(Strong normalization is well-foundedness of the converse relation.)
The ``standard'' constructive definition does not even allow one to show that SN implies WN, see Example \ref{ex:undec}.  We therefore looked at a number of variations of this notion, their classical counterparts all being logically equivalent.  This gave rise to a rich set of concepts, see Figure \ref{f:wf}.

Along the way, we found marginal improvements to a number of standard theorems.
For example, the termination assumption of the famous Newman's Lemma can be slightly weakened....

There are also large classes of ARSs where some needed classical principles are simply valid.  For example, finitely branching relations encompass most ARSs resulting from first-order term rewrite systems.  For such relations, the implication SN to WN requires nothing else beyond plain decidability of
the relation itself $Rxy \lor \lnot Rxy$.

The main conclusion of our work is that most of basic ARS results can be made completely effective, at least for the most basic instance of first-order TRSs.
(This also includes the usual lambda calculus.)

For more exotic rewrite systems, like $\lambda \bot$ or coinductive rewriting,
these decidability assumptions no longer hold, and the utility of ARS results diminishes proportionally.

Finally, our Agda implementation provides a standard ``entry point'' for using these results in formal language development.

@@

META
This is a report about the work that was completed.
Putting ARS on a constructive foundation -- stress that.
